Dirace: Practical Proof Automation of Dirac Notation Equations

### 2025-01-24 main.tex
2025/1/24

\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
\usetikzlibrary{trees}

@@

\section{Motivation and Preliminary}
One interesting fact about the quantum world is that for two maximally entangled states, applying one operator $M$ on one subsystem is equivalent to applying $M^T$, the transpose of $M$, on the other. This property holds no matter how far the two systems are separated, and is expressed as an equation of Dirac notations.
\begin{example}
    \label{ex: motivating}
    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
    \[
    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
    \]
\end{example}
Here $q$ and $r$ are called labels, denoting the system for Dirac notations.
To reason about such equations using computers, we need to formalize the language and work in a proof system.
The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.

@@

With the concrete basis, Dirac notations can be interpreted as matrices. For example, the ket of two basis of qubit states are column vectors $\ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix}$ and $\ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix}$, and the corresponding bras are row vectors in the dual space $\bra{0} = \begin{bmatrix} 1 & 0 \end{bmatrix}$ and $\bra{1} = \begin{bmatrix} 0 & 1 \end{bmatrix}$.
The calculation between Dirac notations then corresponds to matrix operatoins. For example, inner product $\braket{0| 1} = \begin{bmatrix} 0 & 1 \end{bmatrix} \begin{bmatrix} 1 \\ 0 \end{bmatrix} = 0$, and tensor product $\ket{0} \otimes \ket{1} = \ket{(0, 1)} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \otimes \begin{bmatrix} 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \end{bmatrix}$.

@@

\subsubsection{Universal algebra}
We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
A universal algebra defines a signature of function symbols.
Terms in the algebra are constructed by constants, variables or function application on terms. 
Other basic concepts like substitution of variables or pattern matching are also defined.
In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
Reasonings of Dirac notation terms follow the equational logic, which defines an equaitonal relation $e_1 \doteq e_2$ compatible with substitution and term construction:
\[
    \frac{E \vdash s \doteq t}{E \vdash v(s) \doteq v(t)}
    \qquad
    \frac{E \vdash s_1 \doteq t_1 \qquad \cdots \qquad E \vdash s_n \doteq t_n}{E \vdash f(s_1, \dots, s_n) \doteq f(t_1, \dots, t_n)}
\]
This logic essentially formalizes the common sense of equivalence in algebra.

@@

For this purpose, our language involves three layers: the index, the type and the term.
Terms describe the concrete instances like ket, bra and operator, which will be checked and typed.
Index represents classical datatypes, and they appear in type expressions to distinguish between different Hilbert spaces and sets.

\begin{definition}[index syntax]
    The syntax for type indices is
    \begin{align*}
        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
    \end{align*}
\end{definition}
Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type for tensor product space or Cartesian product sets.
% $\mathsf{bool}$ is a constant index with two basis, and is used in Dirac notations of qubit systems.

@@

\begin{definition}[term syntax]
    The syntax for Dirac notation terms is
    \begin{align*}
        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
    \end{align*}
\end{definition}

The generators are explained in order.
$\lambda x : T.e$ is the abstraction for normal functions, and $\mu x.e$ is the abstraction for index functions.
$e_1\ e_2$ is function application. 
% $e_1 \circ e_2$ is the composition operator of Dirac notations.
% $\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
$(e_1, e_2)$ is the basis pair for product types. $0$, $1$, $\ADDS$, $e_1 \times \cdots \times e_n$ and $e^*$ are symbols for scalars. 
The next line includes constant symbols for ket, bra and operator.
$e^\dagger$ denotes the conjugate transpose of $e$. $e_1.e_2$ denotes scaling the term $e_2$ with scalar $e_1$.
$\mathbf{U}(e)$ is the universal set with index $e$. $e_1 \star e_2$ is the Cartesian product. $\sum_{e_1} e_2$ is the big operator sum, which is modelled by folding the function $e_2$ over value sets $e_1$. Usually, the sum body is specified by an abstraction. Therefore we use the notation $\sum_{x \in s} X$ to denote $\sum_{s} \lambda x : T . X$ as well.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. 
% They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator. We encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.

@@

Typing is the procedure to classify the term through a proof system in a typing context. 
Our context consists of the environment $E$ for definitions and assumptions, and the context $\Gamma$ for bound variables. 
$E$ and $\Gamma$ are sequences of assumptions $x : T$ or definitions $x := t : T$.

\begin{definition}[environment and context]
    \begin{align*}
        E ::=\ & [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
    \end{align*}
\end{definition}
Definitions refer to the symbols that can be unfolded. Definition symbols usually have abstract meanings, and here we use them to encode more operations in Dirac notations, such as transpose and trace.
Assumptions are the declarations of the type for variables. In the scenario of equational proofs, variable assumptions implicitly have the universal quantifier.

@@

The typing for functions and applications follow the common practice. 
Take the index function $\mu x. t$ as an example: here $x$ is a bound variable typed as $\Index$, and the type $U\{x/u\}$ of application $(t\ u)$ is obtained by replacing $x$ with the index instance $u$.

\begin{gather*}
    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
    \qquad
    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} \\
    \\
    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
    \qquad
    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
\end{gather*}

@@

\begin{lemma}
    The typing of expressions are decidable and unique.
\end{lemma}

\begin{proof}
    The type can be calculated in the recursive manner. There is at most one typing rule for the same function symbol and argument types, which leads to the uniqueness of typing.
\end{proof}

@@

As is mentioned in the introduction, a subset $E$ of the axioms is not decided by term rewriting. The set consists of:

\begin{align*}
    & \textrm{AC-equivalence} 
    && \textrm{e.g.} \qquad X + Y = Y + X \qquad (X + Y) + Z = X + (Y + Z) \\
    & \textrm{$\alpha$-equivalence}
    && \lambda x . A = \lambda y . A\{x/y\} \\
    & \textrm{SUM-SWAP}
    && \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A \\
    & \textrm{scalar theories} 
    && \textrm{e.g.} \qquad a + 0 = a \qquad a \times (b + c) = a \times b + a \times c
\end{align*}

The equational theories for scalars are treated as a moduloe, and are not considered here. In the implementation, we use the Mathematica kernel to decide scalar equivalences.

The equational axioms provide an operable theory for the proof automation algorithm,
and the denotational semantics can be considered as one model for the theory.
In this sense, the equivalence derived by axioms always imply the equivalence in interpretations.

\begin{lemma}
    For all well-formed context $E[\Gamma]$ and terms $e_1$, $e_2$, if $AX \vdash e_1 = e_2$, then $\sem{e_1} = \sem{e_2}$.
\end{lemma}
\begin{proof}
    Direct check all cases.
\end{proof}

@@

We use the term rewriting technique to decide the equivalence of Dirac notation implied by its axiomatic semantics.
Term rewriting decides equivalence by normalizing the two terms, checking whether they have the same syntax.
The normalization is powered by matching and substitution in universal algebra, using a set of rewriting rules.
As

% CONTINUE HERE

The following two sections discuss how to decide the equivalence of Dirac notations implied by the axioms.
In the previous work, the equational axioms are separated into two parts: a set $E$ of equations that can not be decided by rewriting, and the remaining equations to be decided by term rewriting.
Here, the equational theories $E$ inlude:
\begin{itemize}
    \item commutativity of symbols $a + b$, $a\times b$ and $\delta_{s, t}$,
    \item $\alpha$-equivalence of bound variables, i.e., $\lambda x.t = \lambda y . t\{x/y\}$,
    \item swapping successive summations, i.e., $\sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1}A$, and
    \item equational theories for scalars.
\end{itemize}

In this work, the general idea is to carry through the normalization procedure, so that semantical equivalence can be directly checked by the syntax of normal forms. The procedure of the normalization is displayed in~\Cref{fig: normalization}.

\input{flowchart.tex}

The last three steps deals with the equational theory $E$, while the first three steps use term rewriting to work on the structure of Dirac notations.

\subsection{Normalization modulo $E$ by Term Rewriting}
Term rewriting rules, written as $l \to r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$. The procedure terminates when no more rewritings can be made, and the order of rewritings will be irrelevant if the term rewriting system is \textit{confluent}, which is a desirable property.

To express the term rewriting system, the previous work adheres to use the naive universal algebra, where each function symbol has a fixed arity, and the left hand side pattern only allows constant symbols and variables. This constraint enables them to check the confluence and termination of their term rewriting system with other tools.

% Here to enable more efficient algorithms, our langauge uses AC symbols with a variable arity.

The full list of rewriting rules are in~\Cref{sec: rewriting rules}. Here we present some of them to illustrate the design idea.
The whole rule list consists of several types.
Firstly, we use the following flattening rule to decide the associative with AC symbols.

\[
a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n
\ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n,
\]

and commutativity is left for a sorting algorithm later on.

Some equational axioms imply obvious rewriting diraction, for example:

\begin{align*}
    &\textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce \delta_{s, t} \\
    &\textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\ 1 \\
    &\textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce \mathbf{0}_{\KType(\sigma)}
\end{align*}

The (R-DOT6) says that inner product of two basis will be evaluated to a delta expression,
and (R-DELTA0) further transforms delta on identical basis to scalar $1$.
(R-MULK1) corresponds to the axiom that multiplication on zero vectors results in zero. It has a condition on typing, which is calculated during rewriting.
They share the common feature that right hand side is simpler than the left hand side.

Other part of rules assign the rewriting diraction to equations according to the property of symbols and intuitive observations of their relations.
When different rules rewrites the same term to different forms, rules for confluence purpose are added.

\begin{align*}
    %
    & \text{(R-SCR1)} 
    && a.(b.X) \ \reduce\ (a \times b).X \\
    %
    & \text{(R-DOT10)}
    && (B \cdot O) \cdot K \ \reduce\ B \cdot (O \cdot K), \\
    %
    & \text{(R-DOT11)}
    && \bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K, \\
    %
    & \text{(R-MULB10)}
    && \bra{(s, t)} \cdot (O_1 \otimes O_2)\ \reduce\ (\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2),
\end{align*}
The (R-SCR1) rule follows the idea to reduce linear algebra scaling into scalar multiplications,
and (R-MULB10) transform inner product into tensor product. 
As an example of the completetion rule, here the normalization of term $(\bra{(s, t)} \cdot (O_1 \otimes O_2)) \cdot K$ have two rewriting paths: (a). apply (R-MULB10) and get $((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K$, or
(b). first apply (R-DOT10) and sort the term into $\bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K)$, and then apply (R-DOT11) to get the same result. Here (R-DOT11) is for completeness of cases similar to this example.

There are also rules dealing with sum expressions. For example,

\begin{align*}
    & \textrm{(R-SUM-ELIM2)}
    && i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \\
    & && \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\} \\
    & \textrm{(R-SUM-PUSH5)}
    && (\sum_{i \in M} B)\cdot K \ \reduce\ \sum_{i \in M}(B \cdot K)
\end{align*}

The (R-SUM-ELIM2) rule will try to eliminate the delta expression in summations.
And the (R-SUM-PUSH5) rule will lift sum to the outside of inner product.

One important technique revealed in the DiracDec work is the expansion of variables:

\begin{gather*}
    \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot K).\ket{i}}
    \qquad \qquad
    \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}} \\
    \\
    \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}
\end{gather*}
The above three rules are not terminating, and this is why we have the rewriting-expansion-rewriting steps in the algorithm. On the other hand, we discovered that doing the expansion on all variables only once is already sufficient.

\begin{lemma}
    Let $\text{expand}(e)$ indicate the result of expanding all variables in $e$ once.
    For all well-typed term $e$ in $E[\Gamma]$, we have $\text{expand}(\text{expand}(e)) \downarrow = \text{expand}(e)\downarrow$.
\end{lemma}
\begin{proof}
\end{proof}

@@

In the Mathematica implementation of DiracDec, the equational theory $E$ is decided by a unification, which tries to find a substitution of summation bound variables and AC symbol arguments that makes the two expressions syntactically equivalent. The unification iterates through all permutations, and the complexity is factorial in general.

As the first improvement, the $\alpha$-equivalence are checked by de Bruijn index~\cite{deBruijn1972lambda}, where references to bound variable names are replaced by the distance of the lambda abstraction to the variable. For example, the nominal lambda abstraction $\lambda x. x$ is transformed into $\lambda . 1$, while $\lambda x. \lambda y. (x\ (y\ x))$ will is transformed into $\lambda.\lambda. (2\ (1\ 2))$. Therefore, transformation into de Bruijn index is at the last step of the normalization to check the equivalence of terms with different bound variable names.

>>>>

\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
\usetikzlibrary{trees}
\tikzstyle{arrow} = [thick,->,>=stealth]

@@

\section{Motivation and Preliminary}
One interesting fact about the quantum world is that for two maximally entangled states, applying one operator $M$ on one subsystem is equivalent to applying $M^T$, the transpose of $M$, on the other. This property holds no matter how far the two systems are separated, and is expressed as an equation of Dirac notations.
\begin{example}
    \label{ex: motivating}
    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
    \[
    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
    \]
\end{example}
Here $q$ and $r$ are called labels, denoting the system for Dirac notations.
To reason about such equations using computers, we need to formalize the language and work in a proof system.
% The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.

@@

% With the concrete basis, Dirac notations can be interpreted as matrices. For example, the ket of two basis of qubit states are column vectors $\ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix}$ and $\ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix}$, and the corresponding bras are row vectors in the dual space $\bra{0} = \begin{bmatrix} 1 & 0 \end{bmatrix}$ and $\bra{1} = \begin{bmatrix} 0 & 1 \end{bmatrix}$.
% The calculation between Dirac notations then corresponds to matrix operatoins. For example, inner product $\braket{0| 1} = \begin{bmatrix} 0 & 1 \end{bmatrix} \begin{bmatrix} 1 \\ 0 \end{bmatrix} = 0$, and tensor product $\ket{0} \otimes \ket{1} = \ket{(0, 1)} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \otimes \begin{bmatrix} 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \end{bmatrix}$.

@@

\subsubsection{Universal algebra}
We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
A universal algebra defines a signature of function symbols.
Terms in the algebra are constructed by constants, variables or function application on terms. 
Other basic concepts like substitution of variables or pattern matching are also defined.
In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
Reasonings of Dirac notation terms follow the equational logic. It defines an equaitonal relation compatible with substitution and term construction, which essentially formalizes the common sense of equivalence in algebra.

@@

For this purpose, our language involves three layers: the index, the type and the term.
Terms describe the concrete instances like ket, bra and operator, which will be checked and typed.
Index represents classical datatypes, and they appear in type expressions to distinguish between different Hilbert spaces and sets.
\begin{definition}[index syntax]
    The syntax for type indices is
    \begin{align*}
        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
    \end{align*}
\end{definition}
Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type for tensor product space or Cartesian product sets.

@@

\begin{definition}[term syntax]
    The syntax for Dirac notation terms is
    \begin{align*}
        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
    \end{align*}
\end{definition}
The generators are explained in order.
$\lambda x : T.e$ is the abstraction for normal functions, and $\mu x.e$ is the abstraction for index functions.
$e_1\ e_2$ is function application. 
% $e_1 \circ e_2$ is the composition operator of Dirac notations.
% $\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
$(e_1, e_2)$ is the basis pair for product types. $0$, $1$, $\ADDS$, $e_1 \times \cdots \times e_n$ and $e^*$ are symbols for scalars. 
The next line includes constant symbols for ket, bra and operator.
$e^\dagger$ denotes the conjugate transpose of $e$. $e_1.e_2$ denotes scaling the term $e_2$ with scalar $e_1$.
$\mathbf{U}(e)$ is the universal set with index $e$. $e_1 \star e_2$ is the Cartesian product. $\sum_{e_1} e_2$ is the big operator sum, which is modelled by folding the function $e_2$ over value sets $e_1$. Usually, the sum body is specified by an abstraction. Therefore we use the notation $\sum_{x \in s} X$ to denote $\sum_{s} \lambda x : T . X$ as well.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. 
% They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator. We encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.

@@

Typing is the procedure to classify the term through a proof system in a typing context. 
Our context consists of the environment $E$ for definitions and assumptions, and the context $\Gamma$ for bound variables. 
$E$ and $\Gamma$ are sequences of assumptions $x : T$ or definitions $x := t : T$.
\begin{definition}[environment and context]
    \begin{align*}
        E ::=\ & [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
    \end{align*}
\end{definition}
Definitions refer to the symbols that can be unfolded. Definition symbols usually have abstract meanings, and here we use them to encode more operations in Dirac notations, such as transpose and trace.
Assumptions are the declarations of the type for variables. In the scenario of equational proofs, variable assumptions implicitly have the universal quantifier.

@@

The typing for functions and applications follow the common practice. 
Take the index function $\mu x. t$ as an example: here $x$ is a bound variable typed as $\Index$, and the type $U\{x/u\}$ of application $(t\ u)$ is obtained by replacing $x$ with the index instance $u$.
\begin{gather*}
    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
    \qquad
    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} \\
    \\
    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
    \qquad
    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
\end{gather*}

@@

\begin{lemma}
    The typing of expressions are decidable and unique.
\end{lemma}
\begin{proof}
    The type can be calculated in the recursive manner. There is at most one typing rule for the same function symbol and argument types, which leads to the uniqueness of typing.
\end{proof}

@@

As is mentioned in the introduction, a subset $E$ of the axioms is not decided by term rewriting. The set consists of:
\begin{align*}
    & \textrm{AC-equivalence} 
    && \textrm{e.g.} \qquad X + Y = Y + X \qquad (X + Y) + Z = X + (Y + Z) \\
    & \textrm{$\alpha$-equivalence}
    && \lambda x . A = \lambda y . A\{x/y\} \\
    & \textrm{SUM-SWAP}
    && \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A \\
    & \textrm{scalar theories} 
    && \textrm{e.g.} \qquad a + 0 = a \qquad a \times (b + c) = a \times b + a \times c
\end{align*}
The equational theories for scalars are treated as a moduloe, and are not considered here. In the implementation, we use the Mathematica kernel to decide scalar equivalences.

The equational axioms provide an operable theory for the proof automation algorithm,
and the denotational semantics can be considered as one model for the theory.
In this sense, the equivalence derived by axioms always imply the equivalence in interpretations.
\begin{lemma}
    For all well-formed context $E[\Gamma]$ and terms $e_1$, $e_2$, if $AX \vdash e_1 = e_2$, then $\sem{e_1} = \sem{e_2}$.
\end{lemma}
\begin{proof}
    Direct check all cases.
\end{proof}

@@

We use the term rewriting technique to decide the equivalence of Dirac notation implied by its axiomatic semantics.
Term rewriting decides equivalence by normalizing the two terms, and checking whether they have the same syntax.
The normalization is powered by matching and substitution in universal algebra, using a set of rewriting rules.
As is mentioned in the sematnics, the axioms in $E$ cannot be decided by term rewriting.
DiracDec decides $E$ by examining all possible permutations in the rewriting result.
In this work, the general idea is to carry through the normalization procedure using sorting algorithms. 
The procedure of the normalization is displayed below. 
The first three steps use term rewriting to work on the structure of Dirac notations, while the last three steps deals with the equational theory $E$.
\begin{enumerate}
    \item \textbf{first rewritings}: expand definitions and simplify,
    \item \textbf{variable expansion}: consider scalars expressions for completeness,
    \item \textbf{second rewritings}: normalize modulo $E$,
    \item \textbf{sorting without bound variables}: normalize AC-equivalence,
    \item \textbf{swap successive summation}: normalize SUM-SWAP,
    \item \textbf{de Bruijn normalization}: normalize $\alpha$-equivalence.
\end{enumerate}

% \input{flowchart.tex}

\subsection{Normalization modulo $E$ by Term Rewriting}
Term rewriting rules, written as $l \to r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$. The procedure terminates when no more rewritings can be made.
The full list of rewriting rules are in~\Cref{sec: rewriting rules}, and we present some of them to illustrate the design idea.

Firstly, we use the following flattening rule to decide the associative with AC symbols, because we allow functions with indefinite arities.
\[
a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n
\ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n
\]
Commutativity is left for the sorting algorithm later on. 
Many rewriting rules are obtained by assigning directions to the equational axioms, for example:
\begin{align*}
    &\textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce \delta_{s, t} \\
    &\textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\ 1 \\
    &\textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce \mathbf{0}_{\KType(\sigma)} \\
    &\textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2)\ \reduce\ (O_1 \cdot K_1) \otimes (O_2 \cdot K_2)
\end{align*}
Some of them are obvious. The (R-DOT6) says that inner product of two basis will be evaluated to a delta expression,
and (R-DELTA0) further transforms delta on identical basis to scalar $1$.
(R-MULK1) corresponds to the axiom that multiplication on zero vectors results in zero. It has a condition on typing, which is calculated during rewriting.
Some rules require the intuition. For example, (R-MULK11) shows that we prefer tensor product than multiplication.

As a reference, the term rewriting system for DiracDec is proved complete for the axioms, except the sum symbol. The result is obtained by checking confluence and termination of the system. Our rewriting rules are translations from DiracDec into the typed and abstract language, therefore completeness for corresponding symbols is also expected to hold.

%%%%%%%%%%% It's better to avoid mentioning completion rules.
% Other part of rules assign the rewriting diraction to equations according to the property of symbols and intuitive observations of their relations.
% When different rules rewrites the same term to different forms, rules for confluence purpose are added.
% \begin{align*}
%     %
%     & \text{(R-SCR1)} 
%     && a.(b.X) \ \reduce\ (a \times b).X \\
%     %
%     & \text{(R-DOT10)}
%     && (B \cdot O) \cdot K \ \reduce\ B \cdot (O \cdot K), \\
%     %
%     & \text{(R-DOT11)}
%     && \bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K, \\
%     %
%     & \text{(R-MULB10)}
%     && \bra{(s, t)} \cdot (O_1 \otimes O_2)\ \reduce\ (\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2),
% \end{align*}
% The (R-SCR1) rule follows the idea to reduce linear algebra scaling into scalar multiplications,
% and (R-MULB10) transform inner product into tensor product. 
% As an example of the completetion rule, here the normalization of term $(\bra{(s, t)} \cdot (O_1 \otimes O_2)) \cdot K$ have two rewriting paths: (a). apply (R-MULB10) and get $((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K$, or
% (b). first apply (R-DOT10) and sort the term into $\bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K)$, and then apply (R-DOT11) to get the same result. Here (R-DOT11) is for completeness of cases similar to this example.

We have other rules dealing with big operator sum. For example, 
\begin{align*}
    & \textrm{(R-SUM-ELIM2)}
    && i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \\
    & && \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\} \\
    & \textrm{(R-SUM-PUSH5)}
    && (\sum_{i \in M} B)\cdot K \ \reduce\ \sum_{i \in M}(B \cdot K)
\end{align*}
The (R-SUM-ELIM2) rule will try to eliminate the delta expression in summations.
And the (R-SUM-PUSH5) rule will lift sum to the outside of inner product.
There are no completeness guarantee, but the rules work well in practice.

One technique revealed in the DiracDec work is the expansion of variables, critical for proofs with sum:
\begin{gather*}
    \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot K).\ket{i}}
    \qquad \qquad
    \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}} \\
    \\
    \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}
\end{gather*}
These rules transform variables to the symbolic summation of their decomposition on the basis.
The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}. On the other hand, we discovered that doing the expansion on all variables only once is already sufficient.
\begin{lemma}
    Let $\textrm{expand}(e)$ indicate the result of expanding all variables in $e$ once.
    For all well-typed term $e$ in $E[\Gamma]$, $\textrm{expand}(\textrm{expand}(e))$ and $\textrm{expand}(e)$ have the same normal form.
\end{lemma}
\begin{proof}
    If we expand a ket variable twice, the structure will be reduced to
    \[
        \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot \sum_{j \in \mathbf{U}(\sigma)}(\bra{j}\cdot K).\ket{j}).\ket{i}
        \ \reduce\ 
        \sum_{i \in \mathbf{U}(\sigma)}\sum_{j \in \mathbf{U}(\sigma)}(\bra{j}\cdot K \times \braket{i|j}).\ket{i},
    \]
    the elimination rule of delta symbol returns the term to the one-expansion result. Similar for bra and operator.
\end{proof}

@@

In the Mathematica implementation of DiracDec, the equational theory $E$ is decided by a unification, which tries to find a substitution of summation bound variables that makes the two expressions syntactically equivalent. The unification iterates through all permutations of AC symbol arguments, and the complexity is factorial in the maximum number of AC symbol arguments.

As the first improvement, we check the $\alpha$-equivalence by de Bruijn index~\cite{deBruijn1972lambda}, where references to bound variable names are replaced by the distance of the lambda abstraction to the variable. For example, the nominal lambda abstraction $\lambda x. x$ is transformed into $\lambda . 0$, while $\lambda x. \lambda y. (x\ (y\ x))$ will is transformed into $\lambda.\lambda. (1\ (0\ 1))$. 
Transformation into de Bruijn index is at the last step of the normalization to check the equivalence of terms with different bound variable names.
